---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    warning = FALSE, 
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# tidyFlowCore

<!-- badges: start -->
[![R-CMD-check-bioc](https://github.com/keyes-timothy/tidyFlowCore/workflows/R-CMD-check-bioc/badge.svg)](https://github.com/keyes-timothy/tidyFlowCore/actions)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![Codecov test coverage](https://codecov.io/gh/keyes-timothy/tidyFlowCore/branch/main/graph/badge.svg)](https://app.codecov.io/gh/keyes-timothy/tidyFlowCore?branch=main)
[![GitHub issues](https://img.shields.io/github/issues/keyes-timothy/tidyflowCore)](https://github.com/keyes-timothy/tidyflowCore/issues)
[![GitHub pulls](https://img.shields.io/github/issues-pr/keyes-timothy/tidyflowCore)](https://github.com/keyes-timothy/tidyflowCore/pulls)
<!-- badges: end -->

`tidyFlowCore` is an R package that bridges the gap between flow cytometry analysis using the `flowCore` package and the tidy data principles advocated by the `tidyverse.` It provides a suite of `dplyr`-, `ggplot2`-, and `tidyr`-like verbs specifically designed for working with `flowFrame` and `flowSet` objects as if they were tibbles; however, your data remain `flowCore` `flowFrame`s and `flowSet`s under this layer of abstraction. 

Using this approach, `tidyFlowCore` enables intuitive and streamlined analysis workflows that can leverage both the Bioconductor and tidyverse ecosystems for cytometry data.  

## Installation instructions

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). Then install `tidyFlowCore` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("tidyFlowCore")
```

And the development version from [GitHub](https://github.com/keyes-timothy/tidyflowCore) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("keyes-timothy/tidyFlowCore")
```

## Example

`tidyFlowCore` allows you to treat `flowCore` data structures like tidy `data.frame`s or `tibbles`. It does so by implementing dplyr, tidyr, and ggplot2 verbs that can be deployed directly on the `flowFrame` and `flowSet` S4 classes. 

In this section, we give a brief example of how `tidyFlowCore` can enable a data analysis pipeline to use all the useful functions of the `flowCore` package and many of the functions of the `dplyr`, `tidyr`, and `ggplot2` packages. 

### Load required packages

```{r, warning = FALSE}
library(tidyFlowCore)
library(flowCore)
```



### Read data

```{r example, eval = requireNamespace('tidyFlowCore')}
# read data from the HDCytoData package
bcr_flowset <- HDCytoData::Bodenmiller_BCR_XL_flowSet()
```


### Data transformation

The `flowCore` package natively supports multiple types of data preprocessing and transformations for cytometry data through the use of its `tranform` class. 

For example, if we want to apply the standard arcsinh transformation often used for CyTOF data to our current dataset, we could use the following code: 

```{r}
asinh_transformation <- flowCore::arcsinhTransform(a = 0, b = 1/5, c = 0)
transformation_list <- 
  flowCore::transformList(
    colnames(bcr_flowset), 
    asinh_transformation
  )

transformed_bcr_flowset <- flowCore::transform(bcr_flowset, transformation_list)
```

Alternatively, we can also use the `tidyverse`'s functional programming paradigm to perform the same transformation. For this, we use the mutate-across framework via `tidyFlowCore`:

```{r}
transformed_bcr_flowset <- 
  bcr_flowset |>
  dplyr::mutate(across(-ends_with("_id"), \(.x) asinh(.x / 5)))
```

### Cell type counting

Suppose we're interested in counting the number of cells that belong to each cell type (encoded in the `population_id` column of `bcr_flowset`) in our dataset. Using standard `flowCore` functions, we could perform this calculation in a few steps: 

```{r}
# extract all expression matrices from our flowSet 
combined_matrix <- flowCore::fsApply(bcr_flowset, exprs)

# take out the concatenated population_id column 
combined_population_id <- combined_matrix[, 'population_id']

# perform the calculation
table(combined_population_id)
```

`tidyFlowCore` allows us to perform the same operation simply using the `dplyr` package's `count` function: 

```{r}
bcr_flowset |> 
  dplyr::count(population_id)
```

And `tidyFlowCore` also makes it easy to perform the counting broken down by other variables in our metadata: 

```{r}
bcr_flowset |> 
  # use the .tidyFlowCore_identifier pronoun to access the name of 
  # each experiment in the flowSet 
  dplyr::count(.tidyFlowCore_identifier, population_id)
```


### Nesting and unnesting

`flowFrame` and `flowSet` data objects have a clear relationship with one another in the `flowCore` API - essentially nested `flowFrame`s. In other words, `flowSet`s are made up of multiple `flowFrame`s!

`tidyFlowCore` provides a useful API for converting between `flowSet` and `flowFrame` data structures at various degrees of nesting using the `group`/`nest` and `ungroup`/`unnest` verbs. Note that in the dplyr and tidyr APIs, `group`/`nest` and `ungroup`/`unnest` are **not** synonyms (grouped `data.frames` are different from nested `data.frames`). However, because of how `flowFrame`s and `flowSet`s are structured, `tidyFlowCore`'s `group`/`nest` and `ungroup`/`unnest` functions have identical behavior, respectively.

```{r}
# unnesting a flowSet results in a flowFrame with an additional column, 
# 'tidyFlowCore_name` that identifies cells based on which experiment in the 
# original flowSet they come from
bcr_flowset |> 
  dplyr::ungroup()
```


```{r}
# flowSets can be unnested and renested for various analyses
bcr_flowset |> 
  dplyr::ungroup() |> 
  # group_by cell type
  dplyr::group_by(population_id) |> 
  # calculate the mean HLA-DR expression of each cell population
  dplyr::summarize(mean_expression = mean(`HLA-DR(Yb174)Dd`)) |> 
  dplyr::select(population_id, mean_expression)
```




### Plotting

`tidyFlowCore` also provides a direct interface between `ggplot2` and `flowFrame` or `flowSet` data objects. For example...

```{r}
library(ggplot2)
bcr_flowset |> 
  dplyr::mutate(across(-ends_with("_id"), \(.x) asinh(.x / 5))) |> 
  ggplot(aes(x = `CD45(In115)Dd`, y = `CD4(Nd145)Dd`)) + 
  geom_point(size = 0.1, alpha = 0.1) +
  facet_wrap(facets = vars(population_id))
```


## Citation

Below is the citation output from using `citation('tidyFlowCore')` in R. Please
run this yourself to check for any updates on how to cite __tidyFlowCore__.

```{r 'citation', eval = requireNamespace('tidyFlowCore')}
print(citation('tidyFlowCore'), bibtex = TRUE)
```

Please note that the `tidyFlowCore` was only made possible thanks to many other R and bioinformatics software authors, which are cited either in the vignettes and/or the paper(s) describing this package.

## Code of Conduct

Please note that the `tidyFlowCore` project is released with a [Contributor Code of Conduct](http://bioconductor.org/about/code-of-conduct/). By contributing to this project, you agree to abide by its terms.

## Development tools

* Continuous code testing is possible thanks to [GitHub actions](https://www.tidyverse.org/blog/2020/04/usethis-1-6-0/)  through `r BiocStyle::CRANpkg('usethis')`, `r BiocStyle::CRANpkg('remotes')`, and `r BiocStyle::CRANpkg('rcmdcheck')` customized to use [Bioconductor's docker containers](https://www.bioconductor.org/help/docker/) and `r BiocStyle::Biocpkg('BiocCheck')`.
* Code coverage assessment is possible thanks to [codecov](https://codecov.io/gh) and `r BiocStyle::CRANpkg('covr')`.
* The [documentation website](http://keyes-timothy.github.io/tidyFlowCore) is automatically updated thanks to `r BiocStyle::CRANpkg('pkgdown')`.
* The code is styled automatically thanks to `r BiocStyle::CRANpkg('styler')`.
* The documentation is formatted thanks to `r BiocStyle::CRANpkg('devtools')` and `r BiocStyle::CRANpkg('roxygen2')`.

For more details, check the `dev` directory.

This package was developed using `r BiocStyle::Biocpkg('biocthis')`.


