---
title: "tidyFlowCore"
author: 
  - name: Timothy Keyes
    affiliation:
    - Stanford University School of Medicine
    email: tkeyes@stanford.edu
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('tidyFlowCore')`"
vignette: >
  %\VignetteIndexEntry{tidyFlowCore}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Track vignette build time
startTime <- Sys.time()

## Bib setup
library(RefManageR)

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    tidyverse = citation("dplyr")[1],
    HDCytoData = citation("HDCytoData")[1],
    tidyFlowCore = citation("tidyFlowCore")[1]
)
```

# Basics

## Installing `tidyFlowCore`

`R` is an open-source statistical environment which can be easily modified to enhance its functionality via packages. `r Biocpkg("tidyFlowCore")` is an `R` package available via [Bioconductor](http://bioconductor.org), an open-source repository for R packages related to biostatistics and biocomputing. `R` can be installed on any operating system from [CRAN](https://cran.r-project.org/) after which you can install `r Biocpkg("tidyFlowCore")` by using the following commands in your `R` session:

```{r "install", eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager")
  }

BiocManager::install("tidyFlowCore")

## Check that you have a valid Bioconductor installation
BiocManager::valid()
```

## Preliminaries

`r Biocpkg("tidyFlowCore")` adopts the so-called "tidy" functional programming paradigm developed by Wickham et al. in the `tidyverse` ecosystem of R packages `r Citep(bib[["tidyverse"]])`. For information about the `tidyverse` ecosytem broadly, feel free to reference the (free) [R for Data Science](https://r4ds.hadley.nz/) book, the [tidyverse website](https://www.tidyverse.org/), or [this paper](https://www.biorxiv.org/content/10.1101/2023.09.10.557072v2) describing the larger `tidyomics` project.

`r Biocpkg("tidyFlowCore")` integrates the `flowCore` Bioconductor package's data analysis capabilities with those of the `tidyverse`. If you're relatively unfamiliar with the Bioconductor project, you might be interested in [this blog post](http://lcolladotor.github.io/2014/10/16/startbioc/#.VkOKbq6rRuU).

## Asking for help

Learning to use `R` and `Bioconductor` can be challenging, so it's important to know where to get help. The main place to ask questions about `tidyFlowCore` is the [Bioconductor support site](https://support.bioconductor.org/). Use the `tidyFlowCore` tag there and look at [previous posts](https://support.bioconductor.org/tag/tidyFlowCore/). 

You can also ask questions on GitHub or Twitter. But remember, if you're asking for help, follow the [posting guidelines](http://www.bioconductor.org/help/support/posting-guide/). Make sure to include a simple example that reproduces your issue (a ["reprex"](https://reprex.tidyverse.org/articles/learn-reprex.html)) and your session info to help developers understand and solve your problem.

## Citing `tidyFlowCore`

If you use `r Biocpkg("tidyFlowCore")` for your research, please use the following citation.

```{r "citation"}
citation("tidyFlowCore")
```

# `tidyFlowCore` quick start

`tidyFlowCore` allows you to treat `flowCore` data structures like tidy `data.frame`s or `tibbles`. It does so by implementing dplyr, tidyr, and ggplot2 verbs that can be deployed directly on the `flowFrame` and `flowSet` S4 classes. 

In this section, we give a brief example of how `tidyFlowCore` can enable a data analysis pipeline to use all the useful functions of the `flowCore` package and many of the functions of the `dplyr`, `tidyr`, and `ggplot2` packages. 

## Load required packages

```{r, warning = FALSE}
library(tidyFlowCore)
library(flowCore)
```



## Read data

For our example here, we download some publicly available mass cytometry (CyTOF) data downloadable through the `r Citep(bib[["HDCytoData"]])` package. These data are made available as a `flowCore::flowSet` S4 object, `Bioconductor`'s standard data structure for cytometry data. 

```{r example, eval = requireNamespace('tidyFlowCore')}
# read data from the HDCytoData package
bcr_flowset <- HDCytoData::Bodenmiller_BCR_XL_flowSet()
```

To read more about this dataset, run the following command: 

```{r, eval = FALSE}
?HDCytoData::Bodenmiller_BCR_XL_flowSet
```


## Data transformation

The `flowCore` package natively supports multiple types of data preprocessing and transformations for cytometry data through the use of its `tranform` class. 

For example, if we want to apply the standard arcsinh transformation often used for CyTOF data to our current dataset, we could use the following code: 

```{r}
asinh_transformation <- flowCore::arcsinhTransform(a = 0, b = 1/5, c = 0)
transformation_list <- 
  flowCore::transformList(
    colnames(bcr_flowset), 
    asinh_transformation
  )

transformed_bcr_flowset <- flowCore::transform(bcr_flowset, transformation_list)
```

Alternatively, we can also use the `tidyverse`'s functional programming paradigm to perform the same transformation. For this, we use the mutate-across framework via `tidyFlowCore`:

```{r}
transformed_bcr_flowset <- 
  bcr_flowset |>
  dplyr::mutate(across(-ends_with("_id"), \(.x) asinh(.x / 5)))
```

## Cell type counting

Suppose we're interested in counting the number of cells that belong to each cell type (encoded in the `population_id` column of `bcr_flowset`) in our dataset. Using standard `flowCore` functions, we could perform this calculation in a few steps: 

```{r}
# extract all expression matrices from our flowSet 
combined_matrix <- flowCore::fsApply(bcr_flowset, exprs)

# take out the concatenated population_id column 
combined_population_id <- combined_matrix[, 'population_id']

# perform the calculation
table(combined_population_id)
```

`tidyFlowCore` allows us to perform the same operation simply using the `dplyr` package's `count` function: 

```{r}
bcr_flowset |> 
  dplyr::count(population_id)
```

And `tidyFlowCore` also makes it easy to perform the counting broken down by other variables in our metadata: 

```{r}
bcr_flowset |> 
  # use the .tidyFlowCore_identifier pronoun to access the name of 
  # each experiment in the flowSet 
  dplyr::count(.tidyFlowCore_identifier, population_id)
```


## Nesting and unnesting

`flowFrame` and `flowSet` data objects have a clear relationship with one another in the `flowCore` API - essentially nested `flowFrame`s. In other words, `flowSet`s are made up of multiple `flowFrame`s!

`tidyFlowCore` provides a useful API for converting between `flowSet` and `flowFrame` data structures at various degrees of nesting using the `group`/`nest` and `ungroup`/`unnest` verbs. Note that in the dplyr and tidyr APIs, `group`/`nest` and `ungroup`/`unnest` are **not** synonyms (grouped `data.frames` are different from nested `data.frames`). However, because of how `flowFrame`s and `flowSet`s are structured, `tidyFlowCore`'s `group`/`nest` and `ungroup`/`unnest` functions have identical behavior, respectively.

```{r}
# unnesting a flowSet results in a flowFrame with an additional column, 
# 'tidyFlowCore_name` that identifies cells based on which experiment in the 
# original flowSet they come from
bcr_flowset |> 
  dplyr::ungroup()
```


```{r}
# flowSets can be unnested and renested for various analyses
bcr_flowset |> 
  dplyr::ungroup() |> 
  # group_by cell type
  dplyr::group_by(population_id) |> 
  # calculate the mean HLA-DR expression of each cell population
  dplyr::summarize(mean_expression = mean(`HLA-DR(Yb174)Dd`)) |> 
  dplyr::select(population_id, mean_expression)
```



## Plotting

`tidyFlowCore` also provides a direct interface between `ggplot2` and `flowFrame` or `flowSet` data objects. For example...

```{r}
library(ggplot2)
bcr_flowset |> 
  dplyr::mutate(across(-ends_with("_id"), \(.x) asinh(.x / 5))) |> 
  ggplot(aes(x = `CD45(In115)Dd`, y = `CD4(Nd145)Dd`)) + 
  geom_point(size = 0.1, alpha = 0.1) +
  facet_wrap(facets = vars(population_id))
```



# Reproducibility

The `r Biocpkg("tidyFlowCore")` package `r Citep(bib[["tidyFlowCore"]])` was made possible thanks to the following: 

* R `r Citep(bib[["R"]])`
* `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
* `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])`
* `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`
* `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])`
* `r CRANpkg("sessioninfo")` `r Citep(bib[["sessioninfo"]])`
* `r CRANpkg("testthat")` `r Citep(bib[["testthat"]])`
* `r CRANpkg("tidyverse")` `r Citep(bib[["tidyverse"]])`

This package was developed using `r BiocStyle::Biocpkg("biocthis")`.


Code for creating the vignette

```{r createVignette, eval=FALSE}
## Create the vignette
library("rmarkdown")
system.time(render("tidyFlowCore.Rmd", "BiocStyle::html_document"))

## Extract the R code
library("knitr")
knit("tidyFlowCore.Rmd", tangle = TRUE)
```

Date the vignette was generated.

```{r reproduce1, echo=FALSE}
## Date the vignette was generated
Sys.time()
```

Wallclock time spent generating the vignette.

```{r reproduce2, echo=FALSE}
## Processing time in seconds
totalTime <- diff(c(startTime, Sys.time()))
round(totalTime, digits = 3)
```

`R` session information.

```{r reproduce3, echo=FALSE}
## Session info
library("sessioninfo")
options(width = 120)
session_info()
```



# Bibliography

This vignette was generated using `r Biocpkg("BiocStyle")` `r Citep(bib[["BiocStyle"]])`
with `r CRANpkg("knitr")` `r Citep(bib[["knitr"]])` and `r CRANpkg("rmarkdown")` `r Citep(bib[["rmarkdown"]])` running behind the scenes.

Citations made with `r CRANpkg("RefManageR")` `r Citep(bib[["RefManageR"]])`.

```{r vignetteBiblio, results = "asis", echo = FALSE, warning = FALSE, message = FALSE}
## Print bibliography
PrintBibliography(bib, .opts = list(hyperlink = "to.doc", style = "html"))
```
